Tail Flexibility in the Degrees of Preferential

### 2025-05-07 paper.qmd
make equations & figures larger in qmd

---
title: Tail Flexibility in the Degrees of Preferential
 Attachment Networks
format:
  nature-pdf:
    fig-pos: 'H'
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-basic
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: natbib
    keep-tex: true
    equal-margins: false
    number-sections: true
author:
  - name: Thomas William Boughen
    affiliations:
      - name: Newcastle University
        id: 1
        department: School of Mathematics, Statistics and Physics
  - name: Clement Lee
    affiliations:
      - ref: 1
  - name: Vianey Palacios Ramirez
    affiliations:
      - ref: 1
abstract: |
  Devising the underlying generating mechanism of a real-life network is difficult as, more often than not, only its snapshots are available, but not its full evolution. One candidate for the generating mechanism is preferential attachment which, in its simplest form, results in a degree distribution that follows the power law. Consequently, the growth of real-life networks that roughly display such power-law behaviour is commonly modelled by preferential attachment. However, the validity of the power law has been challenged by the presence of alternatives with comparable performance, as well as the recent findings that the right tail of the degree distribution is often lighter than implied by the body, whilst still being heavy. In this paper, we study a modified version of the model with a flexible preference function that allows super/sub-linear behaviour whilst also guaranteeing that the limiting degree distribution has a heavy tail. We relate the distributions tail heaviness directly to the model parameters, allowing direct inference of the parameters from the degree distribution alone.
keywords: [networks, extremes]
bibliography: refs.bib

@@

## Structure of this paper

@sec-gpa looks at the theoretical limiting behaviour of a GPA model when $m=1$ according to the results from [@rudas07], going on to introduce a flexible class of preference functions that can guarantee a heavy tailed degree distribution while still being flexible in the body. @sec-rec shows that the preference function parameters can be fairly well recovered from the degree distributions of networks simulated from the GPA model using various functions of the class introduced in @sec-gpa. Having shown that for simulated data the model parameters are recoverable from the degree distribution alone, @sec-real builds upon this and attempts to fit the model to degree distributions of real networks; modelling not only the degree distribution itself but also the estimates for the parameters of the preference function assuming the networks evolved according to the GPA scheme. Ultimately, @sec-conc discusses the main results of the paper, pitfalls of this method and future work.

@@

which aligns with the result from [@rudas07], where the expression is derived using the same branching process results.

Using results from [@shimura12] the domain of attraction of this limiting degree distribution can be found.

@@

Previously, in [@krapivsky01], a preference function of the form $k^\alpha +\varepsilon$ has been considered. However, it has been noted that super-linear preferential attachment ($\alpha > 1$) leads to a degenerate limiting degree distribution where at some point one vertex gains the connection from every vertex that joins the network, additionally sub-linear preferential attachment ($\alpha <1$) has been shown to lead to a light tailed limiting degree distribution. Consider a preference function of the form:

@@

$$
\bar F(k) = \begin{cases}
\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k<k_0\\
\left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right)\frac{\Gamma(\lambda^*+k_0^\alpha + \varepsilon)/\beta)}{\Gamma\left((k_0^\alpha + \varepsilon)/\beta\right)} \frac{\Gamma\left(k-k_0 + 1 +\frac{k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(k-k_0 + 1 +\frac{\lambda^* +k_0^\alpha + \varepsilon}{\beta}\right)},&k\ge k_0.
\end{cases}
$${#eq-polysurv}

with $\lambda^*$ satisfying $\hat \rho(\lambda^*)=1$ where:

$$
\hat\rho(\lambda) = \sum_{n=0}^{k_0}\prod_{i=0}^{n-1}\frac{i^\alpha + \varepsilon}{\lambda+i^\alpha + \varepsilon} + \left(\frac{k_0^\alpha + \varepsilon}{\lambda-\beta}\right)\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda + i^\alpha + \varepsilon} 
$$

which must be solved numerically for most parameter choices. Also, note that $\lambda>\beta$ since when $\lambda\le \beta$ the infinite sum no longer converges and instead goes to infinity.

@@

```{r, echo=FALSE, warning=FALSE, cache=TRUE, out.width="80%"}
#| fig-width: 8
#| fig-height: 4
#| fig-cap: Theoretical survival distributions of the limiting degree distributions, according to various combinations of $(\alpha, \beta, \varepsilon)$ and $k_0=20$ of the proposed preferential attachment model.
#| label: fig-polylinsurv

as = c(0.5,1,1.5)
bs = c(0.1,0.5,1,1.5)
eps = c(.01,.1,.5,1)
k0s = 20
x = 0:200
pars = expand.grid(as,bs,eps,k0s,x)
names(pars) = c('a','b','eps','k0','x')
lambdas = numeric(nrow(pars))
for(i in 1:nrow(pars)){
  lambdas[i] = find_lambda2(polylin(pars$a[i], pars$eps[i]), pars$b[i], pars$k0[i])
}
pars = cbind(pars,lambdas)
names(pars) = c('a', 'b', 'eps', 'k0','x','lambda')
surv =numeric(nrow(pars))
for(i in 1:nrow(pars)){
  surv[i] = S(pars$x[i], polylin(pars$a[i], pars$eps[i]),pars$lambda[i], pars$k0[i],pars$b[i])
}
pars = cbind(pars,surv)
names(pars) = c('a', 'b', 'eps', 'k0','x','lambda','surv')

@@

ggplot(data = pars) + geom_line(aes(x=(x+1),y=surv, linetype=as.character(b),colour =as.character(b))) +
  scale_x_log10() + scale_y_log10(limits=c(1e-5,1))+theme(aspect.ratio = 1/2) + theme_bw() + xlab('Total Degree')  +ylab('Survival') + labs(linetype=TeX('\\beta'), colour =TeX('\\beta')) + 
  facet_grid(eps~a,labeller = labeller, scales='free')
```

The survival (@eq-polysurv) can be connected to the IGPD mentioned in [@sec-intro] by using Sterling's approximation of the gamma function to obtain:

@@

$$
\bar F(k) 
\begin{cases}
=\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k<k_0,\\
\approx \left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right) \left(\frac{\beta(k+1-k_0)}{k_0^{\alpha}+\varepsilon} + 1\right)^{-\lambda^*/\beta},&k\ge k_0,
\end{cases}
$$ meaning that for $k\ge k_0$ the limiting degree distribution is similar to $\text{IGPD}\left(\frac{\beta}{\lambda^*}, \frac{k_0^\alpha + \varepsilon}{\lambda^*},k_0-1\right)$.

To assess how close of an approximation this is the theoretical conditional survivals are shown in @fig-approx_surv in colour and their IGPD approximations are shown in grey. The approximation seems to hold up fairly well even for large degrees.

```{r, echo=FALSE, warning=FALSE, cache=TRUE, out.width="80%"}
#| fig-width: 8
#| fig-height: 4
#| fig-cap: Theoretical conditional survivals (grey) alongside their IGPD approximations (coloured).
#| label: fig-approx_surv

@@

```{r, echo=FALSE, warning=FALSE, cache=TRUE, out.width="80%"}
#| fig-width: 12
#| fig-height: 6
#| fig-cap: Heat maps of $\xi$ for various combinations of the parameters of the proposed model.
#| label: fig-polyheat
#| 
rho_optim_ba= Vectorize(function(a,eps, b, k0){
  return(abs(rho(2*b, polylin(a, eps), b, k0)-1))
},vectorize.args = 'a')
find_a_ba = Vectorize(function(b, eps, k0){
  out = optimise(rho_optim_ba, c(0.00001,3), b=b, eps=eps, k0=k0)$minimum
  return(out)
}, vectorize.args = 'b')
N =50
as = seq(0,2,l=N+1)[-1]
bs = seq(0,2,l=N+1)[-1]
eps = c(.01,.1,.5,1)
k0 = c(25,100)
pars = expand.grid(as,bs,eps,k0)
names(pars) = c('a', 'b', 'eps', 'k0')
lambdas = numeric(nrow(pars))
a_for_ba = numeric(nrow(pars))
for(i in 1:nrow(pars)){
  lambdas[i] = find_lambda2(polylin(pars$a[i], pars$eps[i]), pars$b[i], pars$k0[i])
  a_for_ba[i] = find_a_ba(pars$b[i], pars$eps[i], pars$k0[i])
}
pars = cbind(pars,lambdas,a_for_ba)
names(pars) = c('a', 'b', 'eps', 'k0','lambda','ba')
labeller = label_bquote(cols = `epsilon`==.(eps),rows = ~k[0]==.(k0))

ggplot(pars) + geom_raster(aes(x=a,y=b,fill=b/lambda)) +
  geom_line(aes(x=ba, y=b),linetype='dashed', colour='red',lwd=1)+
  scale_fill_paletteer_c(palette='grDevices::Blues',limits=c(0,1),direction=-1)+theme(aspect.ratio = 1)+ylim(min(bs), max(bs))+xlim(min(as), max(as))+
  facet_grid(k0~eps,labeller = labeller) + labs(fill=TeX('\\xi')) + xlab(TeX('\\alpha')) + ylab(TeX('\\beta')) + theme_minimal()

@@

$$
\bar F(k) = \begin{cases}
\frac{\Gamma(\lambda^*+\varepsilon)\Gamma(k+1+\varepsilon)}{\Gamma(\varepsilon)\Gamma(k+1+\lambda^*+\epsilon)},&k<k_0\\
\frac{\Gamma(\lambda^*+\varepsilon)\Gamma(k_0+\varepsilon)}{\Gamma(\varepsilon)\Gamma(k_0+\lambda^*+\varepsilon)}\times \frac{\Gamma\left(\frac{k_0+\varepsilon+\lambda^*}{\beta}\right)}{\Gamma\left(\frac{k_0+\varepsilon}{\beta}\right)}\times\frac{\Gamma\left(k-k_0 + \frac{k_0+\varepsilon}{\beta}+1\right)}{\Gamma\left(k-k_0 + \frac{k_0+\varepsilon+\lambda^*}{\beta}+1\right)},&k\ge k_0
\end{cases}
$$ {#eq-linsurv}

Just like the previous model, the survival can be approximated by:

$$
\bar F(k) \approx \begin{cases}
\frac{\Gamma(\lambda^*+\varepsilon)}{\Gamma(\varepsilon)} (1+\varepsilon)^{\lambda^*}\left(\frac{k}{1+\varepsilon} + 1\right)^{-\lambda^*},&k<k_0\\
\frac{\Gamma(\lambda^*+\varepsilon)\Gamma(k_0+\varepsilon)}{\Gamma(\varepsilon)\Gamma(k_0+\lambda^*+\varepsilon)}\times\left(\frac{\beta(k+1-k_0)}{k_0+\varepsilon} + 1\right)^{-\lambda^*/\beta},&k\ge k_0
\end{cases}
$$

@@

# Recovery {#sec-rec}

The goal of was to provide a possible preference function that is able to explain the growth of real networks with varied tail behaviour in their degree distributions. This section aims to show that the parameters of the model in @sec-model can be recovered from the degree distribution of a network simulated from it.

>>>>

---
title: Tail Flexibility in the Degrees of Preferential
 Attachment Networks
format:
  nature-pdf:
    fig-pos: 'H'
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-basic
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: natbib
    keep-tex: true
    equal-margins: false
    number-sections: true
author:
  - name: Thomas William Boughen
    affiliations:
      - name: Newcastle University
        id: 1
        department: School of Mathematics, Statistics and Physics
  - name: Clement Lee
    affiliations:
      - ref: 1
  - name: Vianey Palacios Ramirez
    affiliations:
      - ref: 1
abstract: |
  Devising the underlying generating mechanism of a real-life network is difficult as, more often than not, only its snapshots are available, but not its full evolution. One candidate for the generating mechanism is preferential attachment which, in its simplest form, results in a degree distribution that follows the power law. Consequently, the growth of real-life networks that roughly display such power-law behaviour is commonly modelled by preferential attachment. However, the validity of the power law has been challenged by the presence of alternatives with comparable performance, as well as the recent findings that the right tail of the degree distribution is often lighter than implied by the body, whilst still being heavy. In this paper, we study a modified version of the model with a flexible preference function that allows super/sub-linear behaviour whilst also guaranteeing that the limiting degree distribution has a heavy tail. We relate the distributions tail heaviness directly to the model parameters, allowing direct inference of the parameters from the degree distribution alone.
keywords: [networks, discrete extremes, degree distribution]
bibliography: refs.bib

@@

The rest of this paper is as follows: @sec-gpa looks at the theoretical limiting behaviour of a GPA model when $m=1$ according to the results from [@rudas07], going on to introduce a flexible class of preference functions that can guarantee a heavy tailed degree distribution while still being flexible in the body. @sec-rec shows that the preference function parameters can be fairly well recovered from the degree distributions of networks simulated from the GPA model using various functions of the class introduced in @sec-gpa. Having shown that for simulated data the model parameters are recoverable from the degree distribution alone, @sec-real builds upon this and attempts to fit the model to degree distributions of real networks; modelling not only the degree distribution itself but also the estimates for the parameters of the preference function assuming the networks evolved according to the GPA scheme. Ultimately, @sec-conc discusses the main results of the paper, pitfalls of this method and future work.

@@

which aligns with the result from @rudas07, where the expression is derived using the same branching process results.

Using results from @shimura12, the domain of attraction of this limiting degree distribution can be found.

@@

Previously, in @krapivsky01, a preference function of the form $k^\alpha +\varepsilon$ has been considered. However, it has been noted that super-linear preferential attachment ($\alpha > 1$) leads to a degenerate limiting degree distribution where at some point one vertex gains the connection from every vertex that joins the network, additionally sub-linear preferential attachment ($\alpha <1$) has been shown to lead to a light tailed limiting degree distribution. Consider a preference function of the form:

@@

$$
\bar F(k) = \begin{cases}
\displaystyle\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k<k_0\\
\displaystyle\left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right)\frac{\Gamma(\lambda^*+k_0^\alpha + \varepsilon)/\beta)}{\Gamma\left((k_0^\alpha + \varepsilon)/\beta\right)} \frac{\Gamma\left(k-k_0 + 1 +\frac{k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(k-k_0 + 1 +\frac{\lambda^* +k_0^\alpha + \varepsilon}{\beta}\right)},&k\ge k_0,
\end{cases}
$${#eq-polysurv}

with $\lambda^*$ satisfying

$$
\hat\rho(\lambda^{*}) = \sum_{n=0}^{k_0}\prod_{i=0}^{n-1}\frac{i^\alpha + \varepsilon}{\lambda^{*}+i^\alpha + \varepsilon} + \left(\frac{k_0^\alpha + \varepsilon}{\lambda^{*}-\beta}\right)\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^{*} + i^\alpha + \varepsilon} = 1,
$$

which has to be solved numerically for most parameter choices. Note that $\lambda>\beta$ since when $\lambda\le \beta$ the infinite sum no longer converges and instead goes to infinity.

@@

```{r, echo=FALSE, warning=FALSE, cache=TRUE, out.width="95%"}
#| fig-width: 8
#| fig-height: 4
#| fig-cap: Theoretical survival distributions of the limiting degree distributions, according to various combinations of $(\alpha, \beta, \varepsilon)$ and $k_0=20$ of the proposed preferential attachment model.
#| label: fig-polylinsurv

as = c(0.5,1,1.5)
bs = c(0.1,0.5,1,1.5)
eps = c(.01,.1,.5,1)
k0s = 20
x = seq(0, 2000, by = 10)
pars = expand.grid(as,bs,eps,k0s,x)
names(pars) = c('a','b','eps','k0','x')
lambdas = numeric(nrow(pars))
for(i in 1:nrow(pars)){
  lambdas[i] = find_lambda2(polylin(pars$a[i], pars$eps[i]), pars$b[i], pars$k0[i])
}
pars = cbind(pars,lambdas)
names(pars) = c('a', 'b', 'eps', 'k0','x','lambda')
surv =numeric(nrow(pars))
for(i in 1:nrow(pars)){
  surv[i] = S(pars$x[i], polylin(pars$a[i], pars$eps[i]),pars$lambda[i], pars$k0[i],pars$b[i])
}
pars = cbind(pars,surv)
names(pars) = c('a', 'b', 'eps', 'k0','x','lambda','surv')

@@

ggplot(data = pars) + geom_line(aes(x=(x+1),y=surv, linetype=as.character(b),colour =as.character(b))) +
  scale_x_log10() + scale_y_log10(limits=c(1e-8,1))+theme(aspect.ratio = 1/2) + theme_bw() + xlab('Total Degree')  +ylab('Survival') + labs(linetype=TeX('\\beta'), colour =TeX('\\beta')) + 
  facet_grid(eps~a,labeller = labeller, scales='free')
```

Applying Stirling's approximation of the gamma function to the second case of @eq-polysurv, we obtain, for $k\geq k_0$,

@@

$$
\bar F(k) 
%\begin{cases}
%=\displaystyle\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k<k_0,\\
\displaystyle\approx \left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right) \left(\frac{\beta(k+1-k_0)}{k_0^{\alpha}+\varepsilon} + 1\right)^{-\lambda^*/\beta},%&k\ge k_0,
%\end{cases}
$$ which is the survival function of the $\text{IGPD}\left(\frac{\beta}{\lambda^*}, \frac{k_0^\alpha + \varepsilon}{\lambda^*},k_0-1\right)$ distribution. This means that, above the threshold of the piecewise preference function, the limiting degree distribution can be approximated by the IGPD with shape parameter $\beta/\lambda^{*}$.

To assess the quality of the approximation, the conditional survivals are shown in @fig-approx_surv in colour and their IGPD approximations are shown in grey. The approximation seems to hold up fairly well even for large degrees.

```{r, echo=FALSE, warning=FALSE, cache=TRUE, out.width="95%"}
#| fig-width: 8
#| fig-height: 4
#| fig-cap: Theoretical conditional survivals (grey) alongside their IGPD approximations (coloured).
#| label: fig-approx_surv

@@

```{r, echo=FALSE, warning=FALSE, cache=TRUE, out.width="95%"}
#| fig-width: 12
#| fig-height: 6
#| fig-cap: Heatmaps of $\xi$ for various combinations of the parameters of the proposed model.
#| label: fig-polyheat
#| 
rho_optim_ba= Vectorize(function(a,eps, b, k0){
  return(abs(rho(2*b, polylin(a, eps), b, k0)-1))
},vectorize.args = 'a')
find_a_ba = Vectorize(function(b, eps, k0){
  out = optimise(rho_optim_ba, c(0.00001,3), b=b, eps=eps, k0=k0)$minimum
  return(out)
}, vectorize.args = 'b')
N =50
as = seq(0,2,l=N+1)[-1]
bs = seq(0,2,l=N+1)[-1]
eps = c(.01,.1,.5,1)
k0 = c(25,100)
pars = expand.grid(as,bs,eps,k0)
names(pars) = c('a', 'b', 'eps', 'k0')
lambdas = numeric(nrow(pars))
a_for_ba = numeric(nrow(pars))
for(i in 1:nrow(pars)){
  lambdas[i] = find_lambda2(polylin(pars$a[i], pars$eps[i]), pars$b[i], pars$k0[i])
  a_for_ba[i] = find_a_ba(pars$b[i], pars$eps[i], pars$k0[i])
}
pars = cbind(pars,lambdas,a_for_ba)
names(pars) = c('a', 'b', 'eps', 'k0','lambda','ba')
labeller = label_bquote(cols = `epsilon`==.(eps),rows = ~k[0]==.(k0))

ggplot(pars) + geom_raster(aes(x=a,y=b,fill=b/lambda)) +
  geom_line(aes(x=ba, y=b),linetype='dashed', colour='red',lwd=1)+
  scale_fill_paletteer_c(palette='grDevices::Blues',limits=c(0,1),direction=-1)+theme(aspect.ratio = 1)+ylim(min(bs), max(bs))+xlim(min(as), max(as))+
  facet_grid(k0~eps,labeller = labeller) + labs(fill=TeX('\\xi')) + xlab(TeX('\\alpha')) + ylab(TeX('\\beta')) + theme_minimal(15)

@@

$$
\bar F(k) = \begin{cases}
\displaystyle\frac{\Gamma(\lambda^*+\varepsilon)\Gamma(k+1+\varepsilon)}{\Gamma(\varepsilon)\Gamma(k+1+\lambda^*+\epsilon)},&k<k_0,\\
\displaystyle\frac{\Gamma(\lambda^*+\varepsilon)\Gamma(k_0+\varepsilon)}{\Gamma(\varepsilon)\Gamma(k_0+\lambda^*+\varepsilon)}\times \frac{\Gamma\left(\frac{k_0+\varepsilon+\lambda^*}{\beta}\right)}{\Gamma\left(\frac{k_0+\varepsilon}{\beta}\right)}\times\frac{\Gamma\left(k-k_0 + \frac{k_0+\varepsilon}{\beta}+1\right)}{\Gamma\left(k-k_0 + \frac{k_0+\varepsilon+\lambda^*}{\beta}+1\right)},&k\ge k_0.
\end{cases}
$$ {#eq-linsurv}

Similar to the general case, the survival function can be approximated by:

$$
\bar F(k) \approx \begin{cases}
\displaystyle\frac{\Gamma(\lambda^*+\varepsilon)}{\Gamma(\varepsilon)} (1+\varepsilon)^{\lambda^*}\left(\frac{k}{1+\varepsilon} + 1\right)^{-\lambda^*},&k<k_0,\\
\displaystyle\frac{\Gamma(\lambda^*+\varepsilon)\Gamma(k_0+\varepsilon)}{\Gamma(\varepsilon)\Gamma(k_0+\lambda^*+\varepsilon)}\times\left(\frac{\beta(k+1-k_0)}{k_0+\varepsilon} + 1\right)^{-\lambda^*/\beta},&k\ge k_0.
\end{cases}
$$

@@

# Simulation study {#sec-rec}

<!-- The goal of was to provide a possible preference function that is able to explain the growth of real networks with varied tail behaviour in their degree distributions. -->
This section aims to show that the parameters of the model in @sec-model can be recovered from the degree distribution of a network simulated from it.
